Finiteness in Cubical Type Theory

### 2020-01-22 paper.tex
more

  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\).
  We define it the standard way, where \(\mathbf{Fin}(0) = \bot\) and
  \(\mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)\).

@@

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{romlemma} \label{discrete-surj}
  \begin{equation}
    \frac{
        A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
      }{
       \text{Discrete}(B) 
      }
  \end{equation}
\end{romlemma}
\begin{romlemma}
  Every split enumerable type is discrete.
\end{romlemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\mathbf{Fin}(n)\) for some \(n\).
  Also, we know that \(\mathbf{Fin}(n)\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
  \qed
\end{proof}
\subsection{Manifest Bishop Finiteness}
\begin{romdefinition}[Manifest Bishop Finiteness]
  \begin{equation} \label{bish-def}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{romdefinition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).
\begin{romdefinition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{romdefinition}

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{romlemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{romlemma}
\begin{romlemma}
  This circle \(S^1\) is Kuratowski finite.
\end{romlemma}
The second of these in particular tells us that the ``Kuratowski-finite types''
are not necessarily sets; it also tells us that we cannot derive decidable
equality from a proof of Kuratowski finiteness.
\section{Relations Between Each Finiteness Definition}
We now look at under what conditions a given finiteness predicate satisfies
another.
\subsection{Split Enumerability and Manifest Bishop Finiteness}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\subsection{Manifest Enumerability and Kuratowski Finiteness}
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\section{Topos}
\begin{itemize}
  \item \(\bot\)
  \item \(\top\)
  \item Bool
  \item \(\Sigma\)
  \item \(\Pi\)
  \item Pre-categories to categories
  \item \(\Pi\)-pretopos
\end{itemize}
\section{Countably Infinite Types}
\begin{itemize}
  \item Definition
  \item \(\Sigma\)
  \item Kleene star
  \item Truncated
  \item Topos
\end{itemize}
\section{Search}
\begin{itemize}
  \item Omniscience
  \item Search
  \item Curry / uncurry
  \item Synth function
\end{itemize}

>>>>

  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\). We
  define it the standard way, where \(\mathbf{Fin}(0) \coloneqq \bot\) and
  \(\mathbf{Fin}(n + 1) &\coloneqq \top \uplus \mathbf{Fin}(n)\).

@@

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{romlemma} \label{discrete-surj}
  \begin{equation}
    \frac{
        A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
      }{
       \text{Discrete}(B) 
      }
  \end{equation}
\end{romlemma}
\begin{romlemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{romlemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\mathbf{Fin}(n)\) for some \(n\).
  Also, we know that \(\mathbf{Fin}(n)\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
  \qed
\end{proof}
\subsection{Manifest Bishop Finiteness}
\begin{romdefinition}[Manifest Bishop Finiteness]
  \begin{equation} \label{bish-def}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{romdefinition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).
\begin{romdefinition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{romdefinition}

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{romlemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{romlemma}
\begin{romlemma}
  This circle \(S^1\) is Kuratowski finite.
\end{romlemma}
The second of these in particular tells us that the ``Kuratowski-finite types''
are not necessarily sets; it also tells us that we cannot derive decidable
equality from a proof of Kuratowski finiteness.
\section{Relations Between Each Finiteness Definition}
We now look at under what conditions a given finiteness predicate satisfies
another.
\subsection{Split Enumerability and Manifest Bishop Finiteness}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both types imply the other.
\begin{romlemma}
  Any manifest Bishop finite type is split enumerable.
\end{romlemma}
\begin{proof}
  The only difference between \(\mathcal{B}\) and \(\mathcal{E}!\) is the
  membership term.
  To derive \(\mathcal{E}!\) from \(\mathcal{B}\), then, we need to derive
  \(\in\) from \(\inunique\).
  We can do this with fst.
  \begin{alignat*}{4}
    &\text{fst} : && \; \Sigma(x : A) , U(x) \rightarrow A \\
    && \simeq      & \; \text{isContr}(A) \rightarrow A \\
    && \simeq      & \; (x \in \mathit{xs}) \rightarrow (x \inunique \mathit{xs})
  \end{alignat*}
  \qed
\end{proof}

To derive \(\mathcal{B}\) from \(\mathcal{E}!\) takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\begin{romlemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\begin{romlemma}
  Any manifestly Bishop finite type is manifestly enumerable.
\end{romlemma}
\begin{romlemma}
  A manifestly enumerable type with decidable equality is manifest Bishop
  finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{romlemma}
  Any manifest Bishop finite type is cardinal finite.
\end{romlemma}
\begin{romtheorem}
  Any cardinal finite type with a total order is Bishop finite.
\end{romtheorem}
\subsection{Manifest Enumerability and Kuratowski Finiteness}
\begin{romlemma}
  Any manifest enumerable type is Kuratowski finite.
\end{romlemma}
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{romtheorem} \label{kuratowski-cardinal}
  \begin{equation}
    \mathcal{K}^f(A) \times \text{Discrete}(A) \simeq \mathcal{C}(A)
  \end{equation}
\end{romtheorem}
\section{Topos}
Our main result in this section is the formal proof that decidable Kuratowski
finite types form a \(Pi\)-pretopos.
As we saw in theorem~\ref{kuratowski-cardinal}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  From lemma~\ref{bishop-equiv} we can prove the goal by constructing a term of
  the type:
  \begin{equation}
    \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq \bot)
  \end{equation}
  To do so, we take \(n\) to be 0, and by definition~\ref{Fin},
  \(\mathbf{Fin}(0) = \bot\).
  The equivalence is simply the identity equivalence.
\end{proof}
\begin{romlemma}
  \(\top\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  Again, we need only provide a suitable cardinality (in this case 1), and an
  equivalence with \(\mathbf{Fin}(1)\) to prove our goal.
  \begin{align*}
    \top &\simeq \mathbf{Fin}(1) \\
          &\simeq \top \uplus \mathbf{Fin}(0) \\
          &\simeq \top \uplus \bot \\
          &\simeq \top
  \end{align*}
  \qed
\end{proof}
\begin{romlemma}
  \(\mathbf{Bool}\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  We prove this by showing an isomorphism with \(\mathbf{Fin}(2)\).
  \begin{align*}
    \text{false} & \leftrightarrow 0 \\
    \text{true} & \leftrightarrow 1
  \end{align*}
  \qed
\end{proof}

\begin{romlemma}
  Split enumerability is closed under \(\Sigma\),
\end{romlemma}
\begin{romlemma}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{romlemma}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

  First, we will simplify things slightly by working only with split
  enumerability.
  As this is equal in strength to manifest Bishop finiteness, any closure proofs
  carry over.

  Secondly, we will replace \(A\) in all places with \(\mathbf{Fin}(n)\).
  Since we have already seen an equivalence between these two types, we are
  permitted to transport along these lines.
  This is the first transport operation.

  The bulk of the proof now is concerned with proving the following:
  \begin{equation}
    \left( \Pi {(x : \mathbf{Fin}(n))} , \mathcal{E}!(A(x)) \right) \rightarrow \mathcal{E}!\left( \Pi {(x : \mathbf{Fin}(n))} , A(x) \right)
  \end{equation}
  Our strategy to accomplish this will be to consider functions from
  \(\mathbf{Fin}(n)\) as \(n\)-tuples over some type family \(T : \mathbb{N}
  \rightarrow \text{Type}\).
  \begin{equation}
    \begin{aligned}
      \mathbf{Tuple}(T, 0)   &= \top \\
      \mathbf{Tuple}(T, n+1) &= T(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{aligned}
  \end{equation}
  This type is manifestly Bishop finite, as it is constructed only from products
  and the unit type.

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\subsection{The Category of Finite Sets}
\subsection{The \(\Pi\)-pretopos of Finite Sets}
\section{Countably Infinite Types}
\begin{itemize}
  \item Definition
  \item \(\Sigma\)
  \item Kleene star
  \item Truncated
  \item Topos
\end{itemize}
\section{Search}
\begin{itemize}
  \item Omniscience
  \item Search
  \item Curry / uncurry
  \item Synth function
\end{itemize}
